---
title: "Longitudinal sampling with SEEPS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Longitudinal sampling with SEEPS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r setup}
library(SEEPS)
set.seed(1945)
```

In this vingette, we address how to perform longitudinal sampling.
In most applications, data is not taken from a single time point, but rather taken over investigations and multiple years.

We'll look at how to implement each of these components in SEEPS.

## Simulation

The basic simulation function `gen_transmission_history_exponential_constant` generates a transmission history for a single population. However, the act of sampling a population removes a set of individuals from a population. Usually, we assume that sampling an individual would effectively remove them from the population (from the large-population view). To properly account for this, we need to use the more detailed API for designing and implementing the simulation.

### The simulation API

The lower level simulation API is concentrated on the `step` function. Users seeking to develop new simulation methods should concentrate on using this function.

The `step` function takes in a state at integer time `t` with a list of parameters, and returns the state at time `t+1`. This setup is designed to be comfortable to users familiar with dynamical systems which often express $x(t_{n+1}) = f(x(t_n); \theta)$.

The `wrap_parameters` function is provided to encapsulate simulation parameters into a named list. Given the list of parameters, the `initialize` method generates an initial state `x0` with introduced one active infection.

SEEPS provides two additional functions to help with simulations that can perform multiple steps.
* `gen_const_phase` performs a fixed number of steps and returns the final state.
* `gen_exp_phase` performs steps until the the number of active cases is at least the `minimum_population` (a parameter)  and is at least 90% of the `maximum_population_target` (another parameter).

### Removing individuals from the active population

Observing the simulation to draw a sample generally removes the individual from the active population. SEEPS provides the `remove_samples` function to perform this operation. `remove_samples` accepts a state and a vector of sample indices to remove from the simulation. A new state is returned with the samples removed.

```r
state <- SEEPS::remove_samples(state, c(1, 2, 3))
```

## A simple example

We'll start with a simple example. We'll simulate a small population with a a target maximum population size of 20 individuals and sample 5 individuals every 12 months, starting when the population is at least 18 individuals (90% of 20), with $R_0 = 3$.

```{r}
states_sampled <- list()
samples_all <- list()

samples_per_year <- 5
# Discover each contact with 80% probability
contact_tracing_parameter <- 0.8

sim_parameters <- SEEPS::wrap_parameters(
    minimum_population = 6,
    offspring_rate_fn = SEEPS::get_biphasic_HIV_rate(params = list("R0" = 3)),
    maximum_population_target = 20,
    total_steps = 0, # We are going to manually control this so the value will not be used
    spike_root = FALSE
)

state <- SEEPS::initialize(sim_parameters) # Setup
state <- SEEPS::gen_exp_phase(state, sim_parameters) # Run until we have at least 90% of the maximum population
state <- SEEPS::initialize(sim_parameters) # Setup
state <- SEEPS::gen_exp_phase(state, sim_parameters) # Run until we have at least 90% of the maximum population

# Take a first sample at the end of the exponential phase
sample_taken <- SEEPS::contact_traced_uniform_restarts_ids(
    state$active, state$parents, samples_per_year, contact_tracing_parameter
)
# Record the sampling
samples_all[[1]] <- list("curr_step" = state$curr_step, "samples" = sample_taken$samples)
# Remove from the population
state <- SEEPS::remove_samples(state, sample_taken$samples)

# Now simulate forward
for (time_pointer in 1:3) {
  state <- SEEPS::gen_const_phase(state, sim_parameters, num_steps = 12)

    sample_taken <- SEEPS::contact_traced_uniform_restarts_ids(
        state$active, state$parents, samples_per_year, contact_tracing_parameter
    )
    # Record the sampling
    samples_all[[time_pointer + 1]] <- list("curr_step" = state$curr_step, "samples" = sample_taken$samples)
    # Remove from the population
    state <- SEEPS::remove_samples(state, sample_taken$samples)
}

# Now that we have collected all of the sampling data, we can flatten it into a single list

sample_times_ <- list()
sample_ids <- list()
ptr <- 1
for (i in seq_along(samples_all)) {
    sample_times_[[ptr]] <- samples_all[[i]]$curr_step
    sample_ids[[ptr]] <- samples_all[[i]]$samples
    ptr <- ptr + 1
}

# Convert to

genealogy <- SEEPS::reduce_transmission_history_bpb2(
    samples = sample_ids,
    parents = state$parents,
    current_step = sample_times_
)

# Convert to a phylogeny

phylogeny <- SEEPS::geneology_to_phylogeny_bpb(
    transmission_history = genealogy$parents,
    infection_times = genealogy$transmission_times,
    sample_times = genealogy$sample_times,
    a = 5, b = 5,
    leaf_sample_ids = genealogy$transformed_sample_indices
)

# Convert to a newick string for plotting
newick_tree_string <- SEEPS::phylogeny_to_newick(phylogeny$phylogeny, mode = "mean", label_mode = "abs")

# Plot the tree
library(ape)
tree <- drop.tip(read.tree(text = newick_tree_string), "")
plot(tree, cex = 0.5)
axis(1, las = 1, col.axis = "black", col.ticks = "black")
axis(1, las = 1, col.axis = "black", col.ticks = "black")
```




## Reporting the entire history of the simulation

One can also use this approach to record when every individual is removed from the population. Rather than recording a set of samples, we record whenever an individual is removed from the active population.

We define an auxiliary function for simulations.

```{r}
gen_transmission_history_exponential_constant_local <- function(minimum_population, # nolint: object_length_linter
                                                                offspring_rate_fn, maximum_population_target, total_steps,
                                                                spike_root = FALSE) {
    # Top level/user function. Users should start with this function.

    parameters <- SEEPS::wrap_parameters(
        minimum_population, offspring_rate_fn,
        maximum_population_target, total_steps,
        spike_root
    )

    state <- SEEPS::initialize(parameters)
    SEEPS::validate_state(state) # Check that the state is valid
    all_states <- list()
    all_states[[1]] <- state
    state <- gen_exp_phase(state, parameters, all_states = TRUE)
    all_states <- c(all_states, state)

  state <- gen_const_phase(all_states[[length(all_states)]], parameters, num_steps = total_steps, all_states = TRUE)
  all_states <- c(all_states, state)
  return(lapply(all_states, SEEPS::clean_up))
}
```

We'll use a different rate function, where the rate is 50 times higher in the first 3 months than in the remainder of the infection. We set the expected length of the each infection to 30 months (2.5 years).

```{r}
offspring_rate_fn <- get_biphasic_HIV_rate_function(50, 3, 30, params = list("R0" = 1.3))
```

```{r}
# maximum_population_target <- 30
maximum_population_target <- 45
simulator_result <- gen_transmission_history_exponential_constant_local(
    minimum_population = 6,
    offspring_rate_fn = offspring_rate_fn,
    # offspring_rate_fn = SEEPS::get_biphasic_HIV_rate(params = list("R0" = 1.3)),
    # offspring_rate_fn = SEEPS::get_biphasic_HIV_rate(params = list("R0" = 1.2)),
    maximum_population_target = maximum_population_target,
    # total_steps = 48
    total_steps = 60
)

removed_list <- list()
sample_times <- list()
birth_list <- list()
```

```{r}
ptr <- 1
# Don't include the 0 individual as they are not part of the population
# 0 is the root of the first infection.
removed_list[[ptr]] <- c()
sample_times[[ptr]] <- c()
birth_list[[ptr]] <- c(1)
ptr <- 2

for (i in 2:length(simulator_result)) {
    # Get the active nodes at i-1 and i
    active_i <- simulator_result[[i]]$active
    active_i_1 <- simulator_result[[i - 1]]$active

    # Compare the active nodes at i-1 and i
    # If the active node is in i-1 but not i, add it to a "removed" list at position i
    removed_active <- active_i_1[!(active_i_1 %in% active_i)]
    new_active <- active_i[!(active_i %in% active_i_1)]
    if (length(removed_active) == 0 && length(new_active) == 0) {
        next # Nothing to add to the samples list
    }

  # Add to removed list
  removed_list[[ptr]] <- removed_active
  birth_list[[ptr]] <- new_active
  # Record the sample time
  sample_times[[ptr]] <- simulator_result[[i]]$t_end
  ptr <- ptr + 1
}

# Final step - get the active nodes at the last step
active_final <- simulator_result[[length(simulator_result)]]$active
# Store it
removed_list[[ptr]] <- active_final
# Record the sample time
sample_times[[ptr]] <- simulator_result[[length(simulator_result)]]$t_end
```

```{r}
# Now record the transmission history
print(removed_list)
print(sample_times)
full_transmission_history <- reduce_transmission_history_mt(
    parents = simulator_result[[length(simulator_result)]]$parents,
    samples = removed_list,
    current_step = sample_times
)
```

Now we can visualize the entire transmission history.

```{r, fig.height = 12}
string <- SEEPS::phylogeny_to_newick(full_transmission_history$geneology, mode = "mean", label_mode = "abs")
string <- SEEPS::phylogeny_to_newick(full_transmission_history$geneology, mode = "mean", label_mode = "abs")
tree <- read.tree(text = string)
plot(ladderize(tree))
axis(1, las = 1, col.axis = "black", col.ticks = "black")
```

We can also plot out the number of new infections per time step.

```{r}
births_ <- lapply(birth_list, length)
deaths <- lapply(removed_list, length)
# births_ <- births[1:length(births) - 1]  # Don't report the final step
deaths_ <- deaths[1:length(deaths) - 1] # Don't report the final step
sample_times_ <- sample_times[1:length(sample_times) - 1] # Don't report the final step
deaths_ <- deaths[1:length(deaths) - 1] # Don't report the final step
sample_times_ <- sample_times[1:length(sample_times) - 1] # Don't report the final step
deaths_ <- unlist(deaths_)
births_ <- unlist(births_)
print(length(births_))
print(length(deaths_))
plot(seq_along(births_), births_, col = "black", type = "S") # label="Births"
lines(seq_along(deaths_), deaths_, col = "red", type = "S") # label="Deaths"
legend(39, y = NULL, legend = c("Births", "Deaths"), col = c("black", "red"), lty = 1:2, cex = 0.8)
plot(seq_along(births_), births_, col = "black", type = "S") # label="Births"
lines(seq_along(deaths_), deaths_, col = "red", type = "S") # label="Deaths"
legend(39, y = NULL, legend = c("Births", "Deaths"), col = c("black", "red"), lty = 1:2, cex = 0.8)

print(length(births_))
print(length(deaths_))

deltas <- births_ - deaths_
print(length(deltas))
plot(1:length(deltas), cumsum(deltas), ylab = "Number of active infections", xlab = "Time index [steps]", ylim = c(-1, maximum_population_target + 2), type = "S")
plot(1:length(deltas), cumsum(deltas), ylab = "Number of active infections", xlab = "Time index [steps]", ylim = c(-1, maximum_population_target + 2), type = "S")
# Draw a horizontal line at the maximum population target
abline(h = maximum_population_target, col = "red", lty = 2)
```
